Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-26 Manuscript/manuscript.Rmd
edits

@@

## TOTAL NUMBER OF TAD BOUNDARY REGIONS DECREASE AS Hi-C DATA RESOLUTION DECREASES

Expectedly, the total number of TAD boundary regions increased with the increased resolution of Hi-C data irrespectively of cell line (Figure 2 A). At higher resolutions, GM12878 was found to contain more TADs, and thus more TAD boundary regions compared to K562. This can largely be attributed to the increased sequencing depth of the GM12878 library. However, this difference was found to dissipate as resolution decreased, resulting in much more similar numbers of TAD boundary regions between the two cell lines at the 50 kb and 100 kb resolutions respectively.

@@

We then sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were specific to one of the two cell lines (Figure 2 B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically.

@@

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3 C).

@@

<!--
## IMBALANCE RATIOS REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently high with very little variability between chromosomes, regardless of cell line.
-->

@@

| Quantity                 | Inter-Resolution | 95% CI          | P-Value  |
|--------------------------|------------------|-----------------|----------|
| Number of Genomic Bins   | 0.464            | [0.244, 0.690]  | < 0.0001 |
| Number of TADs Called    | 0.627            | [0.426, 0.802]  | < 0.0001 |
| Number of TAD Boundaries | 0.611            | [0.407, 0.791]  | < 0.0001 |
| Imbalanced Ratio         | 0.114            | [-0.058, 0.372] | 0.109    |
-->
<!--
## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS 
-->

## DEVELOPING THREE TYPES OF ASSOCIATIONS BETWEEN TAD BOUNDARIES AND EPIGENOMIC FEATURES

<!-- of TAD boundaries were assessed using functional genomic annotations, derived by ChIP-seq peak algorithms. Specific groups of annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] and included chromatin state segementations, histone modifications, and transcription factor binding sites (TFBS).-->

<!--
Using the Jaccard index, we found that specific sets of these cell line-specific annotations were highly colocalized together on the linear genome (Table 3). For GM12878, the range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. Likewise, for the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?; !!!added implication]. The high Jaccard indices among select sets of genomic annotations indicated the colocalization of annotations known to be enriched at TAD boundaries. This gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space of all annotations while retaining the possibly correlated annotations that could be predictive of TAD boundary formation.
-->

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 2A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 2B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is the least common type of association (Figure 2C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.

>>>>

<!--## TOTAL NUMBER OF TAD BOUNDARY REGIONS DECREASE AS Hi-C DATA RESOLUTION DECREASES-->
Expectedly, the total number of TAD boundary regions increased with the increased resolution of Hi-C data irrespectively of cell line (Figure 2A). At higher resolutions, GM12878 was found to contain more TADs, and thus more TAD boundary regions compared to K562. This can largely be attributed to the increased sequencing depth of the GM12878 library. However, this difference was found to dissipate as resolution decreased, resulting in much more similar numbers of TAD boundary regions between the two cell lines at the 50 kb and 100 kb resolutions respectively.

@@

@@

We first sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically.

@@

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C).

@@

@@

| Quantity                 | Inter-Resolution | 95% CI          | P-Value  |
|--------------------------|------------------|-----------------|----------|
| Number of Genomic Bins   | 0.464            | [0.244, 0.690]  | < 0.0001 |
| Number of TADs Called    | 0.627            | [0.426, 0.802]  | < 0.0001 |
| Number of TAD Boundaries | 0.611            | [0.407, 0.791]  | < 0.0001 |
| Imbalanced Ratio         | 0.114            | [-0.058, 0.372] | 0.109    |
-->

## DEVELOPING THREE TYPES OF PREDICTORS DESCRIBING THE ASSOCIATIONS BETWEEN TAD BOUNDARY REGIONS AND EPIGENOMIC FEATURES

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.
